---
title: "Infectious Disease Bioinoformatics Project "
format:
  html:
    embed-resources: true
editor: visual 
---

# Load Libraries

```{r}
#| label: setup
#| include: true

library(tidyverse)
library(readxl)
library(here)
library(pheatmap)
```

# Data Upload

```{r}
#| label: data-upload-sample-list

# the 53 QC-passed sample IDs
sample_list <- readLines(
  here("results_for_analysis", "metadata", "sample_list_QC_passed.txt")
)

cat("QC-passed samples:", length(sample_list), "\n")
head(sample_list)
```

```{r}
#| label: data-upload-plasmidfinder

plasmid_overview <- read_excel(
  here("results_for_analysis", "plasmidfinder",
       "plasmidfinder_summary_our_data.xlsx"),
  sheet = "Detailed_overview"
)

plasmid_aug_raw <- read_excel(
  here("results_for_analysis", "plasmidfinder",
       "plasmidfinder_summary_augmented.xlsx"),
  sheet = "Detailed_overview"
)

plasmid_overview
```

```{r}
#| label: data-upload-metadata

metadata <- read_tsv(
  here("results_for_analysis", "metadata", "s_aureus_metadata.tsv"),
  show_col_types = FALSE
) |>
  rename(
    sample         = `Sequence file name`,
    date           = `Date of isolation`,
    sample_type    = `Sample type`,
    host           = `Sample source`,
    region         = Region,
    animal_contact = `Human Contact with animals`,
    species        = Species,
    ST_meta        = ST
  )

metadata
```

```{r}
#| label: data-upload-mlst

mlst_df <- read_tsv(
  here("results_for_analysis", "mlst", "mlst_summary.txt"),
  col_names = FALSE,
  show_col_types = FALSE
)

colnames(mlst_df) <- c("sample", "species", "ST",
                       "gene1", "gene2", "gene3",
                       "gene4", "gene5", "gene6", "gene7")

mlst_df <- mlst_df |>
  mutate(
    sample = sample |>
      str_remove("^.*/") |>
      str_remove("\\.fasta$")
  ) |>
  filter(sample %in% sample_list)

mlst_df |> count(ST, sort = TRUE)
```

# **Processing**

```{r}
#| label: process-plasmidfinder

PLASMID_IDENTITY <- 90
PLASMID_COVERAGE <- 60

plasmid_df <- plasmid_overview |>
  rename(
    sample   = Sample,
    plasmid  = Plasmid,
    identity = `%Identity`,
    coverage = Coverage
  ) |>
  filter(!is.na(plasmid)) |>
  filter(sample %in% sample_list) |>
  filter(identity >= PLASMID_IDENTITY, coverage >= PLASMID_COVERAGE) |>
  group_by(sample, plasmid) |>
  slice_max(identity, n = 1, with_ties = FALSE) |>
  ungroup()

cat("Plasmid hits after QC:", nrow(plasmid_df), "\n")
count(plasmid_df, plasmid, sort = TRUE)
```

```{r}
#| label: process-plasmidfinder-augmented

plasmid_aug_df <- plasmid_aug_raw |>
  rename(
    sample   = Sample,
    plasmid  = Plasmid,
    identity = `%Identity`,
    coverage = Coverage
  ) |>
  mutate(ST_aug = str_extract(sample, "^ST[0-9]+") |> str_remove("^ST")) |>
  filter(!is.na(plasmid)) |>
  filter(identity >= PLASMID_IDENTITY, coverage >= PLASMID_COVERAGE) |>
  group_by(sample, plasmid) |>
  slice_max(identity, n = 1, with_ties = FALSE) |>
  ungroup()

cat("Augmented plasmid hits after QC:", nrow(plasmid_aug_df), "\n")
plasmid_aug_df |>
  distinct(sample, ST_aug) |>
  count(ST_aug, sort = TRUE)
```

```{r}
#| label: build-final-strain-data

meta_vs_mlst <- metadata |>
  left_join(mlst_df |> select(sample, ST_mlst = ST), by = "sample") |>
  mutate(
    ST_meta_clean = str_remove(ST_meta, "^ST") |> str_trim(),
    ST_meta_clean = case_when(
      ST_meta_clean == "Unknown" ~ NA_character_,
      ST_meta_clean == "Unkown"  ~ NA_character_,   # metadata typo
      TRUE                       ~ ST_meta_clean
    ),
    ST_mlst_clean = as.character(ST_mlst)
  ) |>
  filter(sample %in% sample_list)

# resolve final ST: MLST priority, metadata fallback
final_strain_data <- meta_vs_mlst |>
  mutate(ST_final = case_when(
    ST_mlst_clean != "-"  ~ ST_mlst_clean,
    !is.na(ST_meta_clean) ~ ST_meta_clean
  )) |>
  select(-c(ST_meta, ST_meta_clean, ST_mlst_clean))

cat("final_strain_data rows:", nrow(final_strain_data), "\n")
count(final_strain_data, ST_final, sort = TRUE)
```

```{r}
#| label: build-annotation

# row annotation: ST, host, region — ordered by ST then host
annot_row <- final_strain_data |>
  mutate(
    ST     = paste0("ST", ST_final),
    region = iconv(region, from = "UTF-8", to = "ASCII//TRANSLIT")
  ) |>
  select(sample, ST, host, region) |>
  arrange(ST, host) |>
  column_to_rownames("sample")

# gap positions between ST groups
gap_pos <- annot_row |>
  mutate(row_n = row_number()) |>
  group_by(ST) |>
  summarise(last = max(row_n), .groups = "drop") |>
  arrange(last) |>
  pull(last) |>
  head(-1)

# ST color palette (auto-scales)
st_levels <- sort(unique(annot_row$ST))
st_cols <- setNames(
  colorRampPalette(c("#378ADD", "#fa0606", "#3B6D11", "#BA7517",
                     "#534AB7", "#0F6E56", "#993556", "#5F5E5A",
                     "#D85A30", "#185FA5"))(length(st_levels)),
  st_levels
)

ann_colors <- list(
  ST = st_cols,
  host = c(
    "Cow"   = "#a0d69c",
    "Human" = "#c9937b",
    "MLSTPUB_avg" = "#ebe270"
  ),
  region = c(
    "Bagamoyo" = "#EEEDFE",
    "Morogoro" = "#AFA9EC",
    "Tanga"    = "#7F77DD",
    "D'Salaam" = "#3C3489"
  )
)

head(annot_row)
gap_pos
```

```{r}
#| label: vir-helper

# build an identity matrix for one database, ordered to match the annotation
build_identity_matrix <- function(long_df, db_name, all_samples, annot) {

  sub <- long_df |> filter(hit_database == db_name)

  mat <- sub |>
    pivot_wider(
      id_cols     = sample,
      names_from  = virulence_factor,
      values_from = identity
    )

  # add samples with no hits in this database
  missing <- setdiff(all_samples, mat$sample)
  if (length(missing) > 0) {
    mat <- bind_rows(mat, tibble(sample = missing))
  }

  mat <- mat |>
    column_to_rownames("sample") |>
    as.matrix()

  # match row order to annotation
  mat[rownames(annot), , drop = FALSE]
}
```

# Matrices + heatmaps

```{r}
#| label: discretize-isolate-matrices

discretize <- function(mat) {
  out <- ifelse(is.na(mat), 0, 1)
  dimnames(out) <- dimnames(mat)
  out
}
```

```{r}
#| label: plasmid-matrix

build_plasmid_matrix <- function(long_df, all_samples, annot) {
  mat <- long_df |>
    pivot_wider(id_cols = sample, names_from = plasmid,
                values_from = identity, values_fn = max)

  missing <- setdiff(all_samples, mat$sample)
  if (length(missing) > 0) {
    mat <- bind_rows(mat, tibble(sample = missing))
  }

  m <- mat |>
    column_to_rownames("sample") |>
    as.matrix()

  m[rownames(annot), , drop = FALSE]
}

mat_plasmid <- build_plasmid_matrix(plasmid_df, sample_list, annot_row)
cat("Plasmid:", nrow(mat_plasmid), "x", ncol(mat_plasmid), "\n")

# binarize to match the AMR/virulence style
mat_plasmid <- discretize(mat_plasmid)

annot_row_plasmid  <- annot_row
ann_colors_plasmid <- ann_colors
```

```{r}
#| label: vir-heatmap-function

draw_vir_heatmap <- function(mat, title) {
  # guard: an empty matrix would crash pheatmap's color scaling
  if (ncol(mat) == 0 || all(is.na(mat))) {
    message("Skipping '", title, "' — no genes passed QC.")
    return(invisible(NULL))
  }
  pheatmap(
    mat,
    annotation_row    = annot_row_vir,
    annotation_colors = ann_colors_vir,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    gaps_row     = gap_pos,
    color        = colorRampPalette(c("#cfe3f5", "#0a4d8c"))(50),
    na_col       = "white",
    border_color = "grey90",
    cellwidth    = 30,
    cellheight   = 10,
    fontsize_row = 7,
    fontsize_col = 10,
    angle_col    = 45,
    legend       = TRUE,
    main         = title
  )
}
```

# Augmented Prevelance

```{r}
#| label: plasmid-augmented-prevalence

MIN_AUG_N <- 5

plasmid_aug_st_n <- plasmid_aug_df |>
  distinct(sample, ST_aug) |>
  count(ST_aug, name = "n_aug")

plasmid_prevalence <- plasmid_aug_df |>
  distinct(sample, ST_aug, plasmid) |>
  count(ST_aug, plasmid, name = "n_present") |>
  left_join(plasmid_aug_st_n, by = "ST_aug") |>
  mutate(prevalence = n_present / n_aug * 100)

build_plasmid_mean_row <- function(prev_df, iso_mat, st) {
  reps  <- colnames(iso_mat)
  n_aug <- plasmid_aug_st_n |> filter(ST_aug == st) |> pull(n_aug)
  if (length(n_aug) == 0) n_aug <- 0

  row_vals <- prev_df |>
    filter(ST_aug == st) |>
    select(plasmid, prevalence) |>
    tibble::deframe()

  vec    <- setNames(rep(0, length(reps)), reps)
  shared <- intersect(names(row_vals), reps)
  vec[shared] <- row_vals[shared] / 100

  matrix(vec, nrow = 1,
         dimnames = list(paste0("mean of ST", st, " (n=", n_aug, ")"), reps))
}

add_plasmid_mean_rows <- function(iso_mat, annot, prev_df, min_n = MIN_AUG_N) {
  st_order <- unique(annot[rownames(iso_mat), "ST"])
  out_rows <- list(); gaps <- integer(0); pos <- 0

  for (st in st_order) {
    st_clean  <- str_remove(st, "^ST")
    iso_names <- intersect(rownames(annot)[annot$ST == st], rownames(iso_mat))

    block <- iso_mat[iso_names, , drop = FALSE]
    out_rows[[length(out_rows) + 1]] <- block
    pos <- pos + nrow(block)

    n_aug <- plasmid_aug_st_n |> filter(ST_aug == st_clean) |> pull(n_aug)
    n_aug <- if (length(n_aug) == 0) 0 else n_aug

    if (n_aug >= min_n) {
      gaps <- c(gaps, pos)
      out_rows[[length(out_rows) + 1]] <- build_plasmid_mean_row(prev_df, iso_mat, st_clean)
      pos  <- pos + 1
      gaps <- c(gaps, pos)
    } else {
      gaps <- c(gaps, pos)
    }
  }

  list(mat = do.call(rbind, out_rows), gaps = head(gaps, -1))
}

plasmid_aug <- add_plasmid_mean_rows(mat_plasmid, annot_row, plasmid_prevalence)

# mark ST152 isolate 0-cells as -1 for yellow highlight (mean row excluded)
st152_isolates <- intersect(rownames(annot_row_plasmid)[annot_row_plasmid$ST == "ST152"],
                            rownames(plasmid_aug$mat))
m <- plasmid_aug$mat
block <- m[st152_isolates, , drop = FALSE]
block[block == 0] <- -1
m[st152_isolates, ] <- block
plasmid_aug$mat <- m

cat("Plasmid matrix with mean rows:", nrow(plasmid_aug$mat), "rows\n")
```

```{r}
#| label: highlight-st152

# flag ST152 ISOLATE rows (not the mean row) and recode their 0s as -1
mark_st152 <- function(aug_mat, annot) {
  st152_isolates <- rownames(annot)[annot$ST == "ST152"]
  st152_isolates <- intersect(st152_isolates, rownames(aug_mat))   # exclude "mean of ST152..."
  
  m <- aug_mat
  block <- m[st152_isolates, , drop = FALSE]
  block[block == 0] <- -1
  m[st152_isolates, ] <- block
  m
}
```

```{r}
#| label: augmented-annotation

make_extended_annot <- function(aug_mat) {
  new_rows <- setdiff(rownames(aug_mat), rownames(annot_row_vir))

  extra <- tibble(
    rowname = new_rows,
    ST      = str_extract(new_rows, "ST[0-9]+"),
    host    = "MLSTPUB_avg",                                   # was NA
    region  = NA_character_
  ) |>
    column_to_rownames("rowname")

  annot_ext <- bind_rows(annot_row_vir, extra)
  annot_ext[rownames(aug_mat), , drop = FALSE]
}
```

```{r}
#| label: plasmid-heatmap
#| fig-width: 10
#| fig-height: 11

make_extended_annot_plasmid <- function(aug_mat) {
  new_rows <- setdiff(rownames(aug_mat), rownames(annot_row_plasmid))

  extra <- tibble(
    rowname = new_rows,
    ST      = str_extract(new_rows, "ST[0-9]+"),
    host    = "MLSTPUB_avg",
    region  = NA_character_
  ) |>
    column_to_rownames("rowname")

  annot_ext <- bind_rows(annot_row_plasmid, extra)
  annot_ext[rownames(aug_mat), , drop = FALSE]
}

annot_plasmid <- make_extended_annot_plasmid(plasmid_aug$mat)

png("plasmid_heatmap.png",
    width = 10, height = 11, units = "in",
    res = 300, bg = "white")

pheatmap(
  plasmid_aug$mat,
  annotation_row    = annot_plasmid,
  annotation_colors = ann_colors_plasmid,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  gaps_row     = plasmid_aug$gaps,
  color        = c("#eeff01",
                   colorRampPalette(c("white", "black"))(100)),
  breaks       = c(-1.5, -0.5, seq(0, 1, length.out = 100)),
  na_col       = "white",
  border_color = "grey90",
  cellwidth    = 30,
  cellheight   = 10,
  fontsize_row = 7,
  fontsize_col = 10,
  angle_col    = 45,
  legend       = TRUE,
  main         = "Plasmid replicon presence/absence + MLSTPUB strain averages"
)

dev.off()
```

\

### What a "rep" is in PlasmidFinder

PlasmidFinder doesn't detect whole plasmids — it detects **replication initiation genes** (`rep` genes). Each plasmid carries a `rep` gene that encodes the protein that starts plasmid replication. A `rep` family is essentially a fingerprint of *which kind* of plasmid an isolate carries. All seven distinct types of replicases found in staphylococcal plasmids are grouped into families like RepL (rep10), Rep_trans (rep7a), Rep_1 (rep13, rep21...), Rep_3 (rep5a), RepA_N (rep15, rep19, rep20...).

### rep10

-   **Family:** RepL

-   **Plasmid type:** small, low-copy-number cryptic plasmids; the canonical example is **pT181-type/pE194-type** plasmids

-   **Resistance cargo it commonly carries:** rep10 carries the erm(C) gene, which gives **macrolide-lincosamide-streptogramin B (MLSB) resistance** — so erythromycin, clindamycin, etc. [[Springer]{.underline}](https://link.springer.com/article/10.1186/s12920-024-01979-4)

-   **Clinical relevance for your project:** if a Tanzanian isolate has rep10, it's likely macrolide-resistant. In the Tanzania MRSA paper, rep10 was found in 72.0% of MRSA isolates — extremely common in that population.

[[PubMed Central]{.underline}](https://pmc.ncbi.nlm.nih.gov/articles/PMC11323609/)[:]{.underline}

-   Methicillin-resistant *Staphylococcus aureus* (MRSA) colonization increases the risk of subsequent infection by MRSA strain complex interlinking between hospital and community-acquired MRSA which increases the chance of drug resistance and severity of the disease.

The "interlinking" the sentence is getting at: **these buckets have stopped being clean.** Over the past 15-20 years, CA-MRSA strains (especially CC8 / USA300) have moved *into* hospitals, and HA-MRSA strains have moved *into* the community. So a CC8 isolate from a hospital patient today might genetically be a USA300 (CC8 CA-MRSA), not the older hospital CC8 lineage.

When this happens, you get the worst of both worlds: a strain with the **virulence of CA-MRSA** + the **multi-drug resistance of HA-MRSA**. That's the "increases chance of drug resistance and severity of disease" part.

-   Genomic characterization of *Staphylococcus aures* strains isolated from patients attending **regional referral hospitals in Tanzania.**

Among 103 sequences of *S. aureus*, 48.5% (50/103) carry the *mecA* gene for MRSA. High proportions of MRSA were observed among participants aged between 18 and 34 years (52.4%), in females (54.3%), and among outpatients (60.5%). The majority of observed MRSA carried plasmids rep5a (92.0%), rep16 (90.0%), rep7c (90.0%), rep15 (82.0%), rep19 (80.0%) and rep10 (72.0%). Among all plasmids observed rep5a, rep16, rep20, and repUS70 carried the *blaZ* gene, rep10 carried the *erm(C)* gene and rep7a carried the *tet(K)* gene. MLST and phylogeny analysis reveal high diversity among MRSA. Six different clones were observed circulating at selected regional hospitals and MRSA with ST8 was dominant.

AMRfinder: Cool, our data does not have mecA at all even though is common

### rep16

-   **Family:** Inc18 (yes, "Inc18" is the database name but rep16 is the gene)

-   **Plasmid type:** broad-host-range conjugative plasmids — these can jump between Gram-positive species (staphylococci, enterococci, streptococci), which makes them important for cross-species AMR spread

-   **Resistance cargo:** rep16 carries the blaZ gene, encoding **β-lactamase** — penicillin resistance. Found on plasmids like pSaaS6159, pSAS, pSJH101 and pWBG759. [[Springer]{.underline}](https://link.springer.com/article/10.1186/s12920-024-01979-4)[[Frontiers]{.underline}](https://www.frontiersin.org/journals/microbiology/articles/10.3389/fmicb.2021.659116/full)

-   **Clinical relevance:** rep16 is one of the most common plasmid types globally in *S. aureus*. In the East China study, rep16 was found in 53.8% of isolates and was the single most prevalent rep type. In Tanzania, 90% of MRSA. [[Wiley Online Library]{.underline}](https://onlinelibrary.wiley.com/doi/10.1155/ijfo/5547073)[[PubMed Central]{.underline}](https://pmc.ncbi.nlm.nih.gov/articles/PMC11323609/)

### rep5a

-   **Family:** Rep_3

-   **Plasmid type:** **pN315-type** plasmids — pN315 is a well-characterized large *S. aureus* plasmid carrying multiple resistance and virulence determinants

-   **Resistance cargo:** rep5a, found on pN315 plasmids, was linked to blaZ (penicillin), qacA (antiseptics) and mupA (mupirocin) genes. So it's β-lactamase + biocide resistance + mupirocin resistance (which matters because mupirocin is used for MRSA decolonization). [[Springer]{.underline}](https://link.springer.com/article/10.1007/s11274-026-04956-4)

-   **Clinical relevance:** another globally dominant type. 92.0% of Tanzanian MRSA carried rep5a. [[PubMed Central]{.underline}](https://pmc.ncbi.nlm.nih.gov/articles/PMC11323609/)
